---
title: 'Homework I: Life Tables'
subtitle: 'Cours: Analyse de Données M1 Mathématiques/M1 Mathématiques-Informatique'
author: "LECHUGA Julian and MORAKHOVSKI Alexander"
output:
  html_notebook: 
    code_folding: none
    number_sections: yes
    toc: yes
---

# Objectives

This notebook aims at

- working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
- visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).# Life dataLife data tables can be downloaded from [https://www.mortality.org](https://www.mortality.org). Some data have been downloaded for you and are available [here]().

See also [https://www.lifetable.de](https://www.lifetable.de).

If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.

```{r}
pacman::p_load(tidyverse)
pacman::p_load(tidymodels)
library(dplyr)
```

## Life table inspection

Load the data corresponding to the country you are supposed to work on.You may use `readr::read_delim()` with appropriate options.

For  country denoted by country code `xxx`, you will load three life tables:

- `xxx.fltper_1x1.txt` where `f` stands for _female_
- `xxx.mltper_1x1.txt` where `m` stands for _male_
- `xxx.bltper_1x1.txt` where `b` stands for _both_

Name the data.frames `data_f`, `data_m`, `data_b`.

```{r}
country_code <- list(fr='FRACNP',
                     be='BEL',
                     gb='GBRTENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')
country <- 'sp'
ww_I <- 1913:1921

#ww_I <- 1938:1946 # (US)
```



```{r}
# TODO:
file_path <- sprintf("LIFE_TABLES/%s", country_code[country][1])
data_f <- readr::read_delim(paste(file_path,".fltper_1x1.txt",sep=""), delim = " ", skip=2)
data_m <- readr::read_delim(paste(file_path,".mltper_1x1.txt",sep=""), delim = " ", skip =2)
data_b <- readr::read_delim(paste(file_path,".bltper_1x1.txt",sep=""), delim = " ", skip=2)
data_f
#data_m
#data_b
data_f[4,]
```

For each life table, each row is identified by `Year` and `Age`.
Read the documentation and concisely explain the meaning of columns

Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns:

`mx`: Central death rate between ages x and x+n. # of Individuals dead / # individuals living in that age group during that year. R = (dx/Lx)

`qx`: Probability of death between ages x and x+n. Probability of dying within 12 months after reaching age x, number of deaths divided by the total amount of individuals. R = (dx/lx)

`ax`: Average length of survival between ages x and x+n for persons dying in the interval. ???

`lx`: Number of survivors at exact age x, assuming l(0) = 100,000. Population of each age group that survived for that year. R = (previous lx-dx) 

`dx`: Number of deaths between ages x and x+n. Number of individuals that died for a specific age group in that year. 
R = (qx*lx)

`ex`: Life expectancy at exact age x (in years). Amount of additional years an individual in a specific age group is expected to live

`Lx`: Number of person-years that lived between ages x and x+n

`Tx`: Number of person-years remaining after exact age x. Number of individuals that are aged >= x. (Previous Tx - Previous Lx)


- What is the class of `data_b, ...`?
- Inspect `data_f` using `glimpse`.

```{r}
# TODO:
class(data_f)
class(data_m)
class(data_b)
glimpse(data_f)
```

Which columns needs retyping?

## Life table cleaning

Remove white spaces from column names. You may find useful functions in package `stringr`.
Try to follow the **DRY** principle (Don't Repeat Yourself).
You need to perform the same operation on every column name.

You may use fonctional programming (using `purrr::` or `plyr::`)
to process all column names of `data_xxx` using one expression.
Some verbs in `dplyr::` provide you with shortcuts.Write a function `retype` that can tidy data.frames
with the same schema as `data_xxx` (cleaning column names if necessary, casting
some columns to `numeric`, others to `double`).

```{r}
retype <- function(data){
  #String cleaning
  names(data)<-str_trim(names(data))
  data<-data%>%
    dplyr::mutate(Year=as.numeric(Year),
                  Age=as.numeric(Age),
                  mx=as.double(mx),
                  qx=as.double(qx),
                  ax=as.double(ax),
                  lx=as.numeric(lx),
                  dx=as.numeric(dx),
                  Lx=as.numeric(Lx),
                  Tx=as.numeric(Tx),
                  ex=as.double(ex))
  data
}
```


Useful tools:

- `dplyr::mutate_at`
- `dplyr::rename`
- `dplyr::rename`Apply function `retype` to the three data.frames.

```{r}
data_f<-retype(data_f)
data_m<-retype(data_m)
data_b<-retype(data_b)
head(data_f)
head(data_m)
head(data_b)
```

# Hazard rates, mortality quotients

Plot mortality quotients (column `mx`) for women as a function of `Age` for  years `1913` up to `1921`.
Use aesthetics to distinguish years. You will need to categorize  the `Year`  column
(`forcats::` may be helpful).

Pay attention to axes labels, to legends. Assess logarithmic scales.

```{r}
data_f %>% # We are taking the categorical variables to change them to a factor and updating the dataset
  dplyr::mutate(Year = forcats::as_factor(Year)) -> data_f
plot(x=data_f$Age[data_f$Year<1921], y =data_f$mx[data_f$Year<1921], col = "red", main = "Women Mortality Quotients", pch=19)
```

Plot mortality quotients (column `mx`) for men as a function of `Age` for  years `1913` up to `1921`.

Note that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
xdata<-data_m$Age[data_m$Year<1921]
ydata<-data_m$mx[data_m$Year<1921]

plot(x=xdata, y=ydata, col = "blue", xlab="Age", ylab="Rate",sub="Years 1913-1921", main = "Men Mortality Quotients", pch=19)
plot(x=data_m$Age[data_m$Year<1921], y =data_m$mx[data_m$Year<1921], col = "blue", main = "Men Mortality Quotients", pch=19)
```

## Facetting

Build a data.frame called `big_data`, where rows are
identified by `Year, Age, Sex` and comes either
from `data_f` or `data_m`. Column `Sex` should be a `factor`
with two levels called `Male` and `Female`.

You may proceed as follows:- Add a character column "Sex" to `data_f` and `data_m`.

  - Stack the two data.frames (using `rbind`) into data.frame named `big_data`.
- Cast  column `Sex` as a factor.

```{r}
data_f$Sex <- forcats::as_factor("Female")
data_m$Sex <- forcats::as_factor("Male")
big_data <- rbind(select(data_f, Year, Age, Sex), select(data_m, Year, Age, Sex))
big_data
```

Plot mortality quotients (column `mx`) as a function of `Age` for
years `1913` up to `1921` facet according to `Sex`.

```{r}
# TODO:
```

Map `Sex` to some aesthetics and facet by `Year`.

```{r}
# TODO:
```

List puzzling features on those plots.
: TODO:

Do you have clues about those puzzling aspects?
: TODO:# Trends## Infant mortality ratePlot mortality quotients at ages $0, 1, 5$  as a function of time.
Distinguish women and men.

```{r}
ages <- c(0, 1, 5)

# TODO:
```

Can you match abrupt changes and historical events?
: TODO:

## As times goes byPlot mortality quotients at ages $15, 20, 40, 60$  as a function of time.
Distinguish women and men.

Note again that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
ages <- c(15, 20, 40, 60)

# TODO:
```

TODO:
: List puzzling features

## Mortality quotients

  - Plot mortality quotient as a function of age for years $1825, 1850, 1875, \ldots, 2000$, facet with respect to `Sex`
- Compare different scales (logarithmic or linear) on the different axes

```{r}
years <- seq(1825, 2000, by = 25)

# TODO:
```

# Rearrangement

From data.frame `data_m, data_f`, extract a numerical matrix `A` where rows correspond to year (say `t`), columns to age (say `x`), and entry
for year `t` and age `x` is the mortality quotient at age `x` for year `t`. Finally, we take logarithms:  $A_{t,x} = \log m_{t,x}$.

You may use functions `pivot_wider`, `pivot_longer` from `tidyr::` package.

```{r}
# TODO:
```

> Factorizations of $A$ play a significant role in demographic modeling. More on this topic in PCA chapter.

# Life expectancy

Write a function that takes as input a vector of mortality quotients, as well as an age, and returns the residual life expectancy corresponding to the vector and the given age.

```{r}
# TODO:
```

```{r}
rex <- function(mx, age){
  # TODO:
}

```

Write a function that takes as input
a _life table_ and returns a matrix $E$ where rows correspond to `Year`
and columns to `Age` with $E_{t,x}$ equal to the residual life expectancy at age $x$ for year $t$.

```{r}
# TODO:
full_rex <- function(lt)
{
  # TODO:
}
```

Plot residual life expectancy as a function of `Year` at ages $60$ and $65$.
You may proceed by first performing aggregation on `data_...` (grouping by `Year`),
and then compute residual life expectancy at different ages
by slicing the `mx` column. Recall that when mapping hazard rate to
residual life expextancy, the order of multiplications matters. How could you secure
your computation so that it still works if the rows of `data_...` are shuffled ?

Function  `tidyr::pivot_longer` may help.

```{r}
# TODO:
```

# References

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
