---
title: 'Homework I: Life Tables'
subtitle: 'Cours: Analyse de Données M1 Mathématiques/M1 Mathématiques-Informatique'
author: "Julian Lechuga and Alexander Morakhovski"
output:
  html_notebook: 
    code_folding: none
    number_sections: yes
    toc: yes
---

# Objectives

This notebook aims at

- working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
- visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).# Life dataLife data tables can be downloaded from [https://www.mortality.org](https://www.mortality.org). Some data have been downloaded for you and are available [here]().

See also [https://www.lifetable.de](https://www.lifetable.de).

If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.

```{r}
pacman::p_load(tidyverse)
pacman::p_load(tidymodels)
library(dplyr)
library(viridis)
```

## Life table inspection

Load the data corresponding to the country you are supposed to work on.You may use `readr::read_delim()` with appropriate options.

For  country denoted by country code `xxx`, you will load three life tables:

- `xxx.fltper_1x1.txt` where `f` stands for _female_
- `xxx.mltper_1x1.txt` where `m` stands for _male_
- `xxx.bltper_1x1.txt` where `b` stands for _both_

Name the data.frames `data_f`, `data_m`, `data_b`.

```{r}
country_code <- list(fr='FRACNP',
                     be='BEL',
                     gb='GBRTENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')

country <- 'fr'

ww_I <- 1913:1921
#ww_I <- 1938:1946 # (US)
```



```{r}
# TODO
file_path <- sprintf("LIFE_TABLES/%s", country_code[country][1])
data_f <- readr::read_delim(paste(file_path,".fltper_1x1.txt",sep=""), delim = " ", skip=2)
data_m <- readr::read_delim(paste(file_path,".mltper_1x1.txt",sep=""), delim = " ", skip =2)
data_b <- readr::read_delim(paste(file_path,".bltper_1x1.txt",sep=""), delim = " ", skip=2)
data_f
#data_m
#data_b
glimpse(data_f)
```

For each life table, each row is identified by `Year` and `Age`.
Read the documentation and concisely explain the meaning of columns

Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns:

`mx`
: TODO:
The number of deaths during the year divided by the
average number alive during the year
`qx`
: TODO:
The number of deaths during the
year divided by the number alive at the beginning of the year.

`ax`
: TODO:
Average length of survival between ages x and x+1 for persons dying in the interval

`lx`
: TODO:
The number of persons surviving to exact age x

`dx`
: TODO:
Number of deaths between ages x and x+1

`ex`:
: TODO:
Life expectancy at exact age x (in years)

- What is the class of `data_b, ...`?
- Inspect `data_f` using `glimpse`.

```{r}
# TODO:
class(data_b)
glimpse(data_f)
```

Which columns needs retyping?
: TODO:
None

## Life table cleaning

Remove white spaces from column names. You may find useful functions in package `stringr`.
Try to follow the **DRY** principle (Don't Repeat Yourself).
You need to perform the same operation on every column name.

You may use fonctional programming (using `purrr::` or `plyr::`)
to process all column names of `data_xxx` using one expression.
Some verbs in `dplyr::` provide you with shortcuts.Write a function `retype` that can tidy data.frames
with the same schema as `data_xxx` (cleaning column names if necessary, casting
some columns to `numeric`, others to `double`).

```{r}
retype <- function(data){
  #String cleaning
  names(data)<-str_trim(names(data))
  data$Age<-str_trim(data$Age)
  data<-data%>%
    dplyr::mutate(Year=as.numeric(Year),
                  Age=forcats::as_factor(Age),
                  mx=as.double(mx),
                  qx=as.double(qx),
                  ax=as.double(ax),
                  lx=as.numeric(lx),
                  dx=as.numeric(dx),
                  Lx=as.numeric(Lx),
                  Tx=as.numeric(Tx),
                  ex=as.double(ex))
  data
}
```

Useful tools:

- `dplyr::mutate_at`
- `dplyr::rename`
- `dplyr::rename`Apply function `retype` to the three data.frames.

```{r}
# TODO:
data_f<-retype(data_f)
data_m<-retype(data_m)
data_b<-retype(data_b)
head(data_f)
head(data_m)
head(data_b)

glimpse(data_f)

```

# Hazard rates, mortality quotients

Plot mortality quotients (column `mx`) for women as a function of `Age` for  years `1913` up to `1921`.
Use aesthetics to distinguish years. You will need to categorize  the `Year`  column
(`forcats::` may be helpful).

Pay attention to axes labels, to legends. Assess logarithmic scales.

```{r}
# TODO:
glimpse(data_f)
data_f %>% 
dplyr::filter(Year %in% ww_I) %>%
ggplot(mapping= aes(colour = Year,x=Age,y=mx)) +  
geom_point()+ xlab("Age")+ ylab("Mortality Quotients") +ggtitle(paste("Mortality Quotients as a function of Age during the First World War"))+scale_x_discrete(breaks=data_f$Age[seq(1,111,by=10)]) + scale_y_log10()+theme_bw()

```

Plot mortality quotients (column `mx`) for men as a function of `Age` for  years `1913` up to `1921`.

Note that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
# TODO:
dplyr::filter(data_m, Year %in% ww_I) -> sub_sample
ggplot(sub_sample, mapping= aes(colour = Year,x=Age,y=mx)) +   
geom_point()+ scale_x_discrete(breaks=data_m$Age[seq(1,111,by=10)])+ scale_y_log10()+theme_bw()

```

## Facetting

Build a data.frame called `big_data`, where rows are
identified by `Year, Age, Sex` and comes either
from `data_f` or `data_m`. Column `Sex` should be a `factor`
with two levels called `Male` and `Female`.

You may proceed as follows:- Add a character column "Sex" to `data_f` and `data_m`.

  - Stack the two data.frames (using `rbind`) into data.frame named `big_data`.
- Cast  column `Sex` as a factor.

```{r}
# TODO:
data_f$Sex='Female'
data_m$Sex='Male'
data_f$Sex <- as.factor(data_f$Sex)
data_m$Sex <- as.factor(data_m$Sex)

big_data<-rbind(data_f, data_m)
glimpse(big_data)
```

Plot mortality quotients (column `mx`) as a function of `Age` for
years `1913` up to `1921` facet according to `Sex`.

```{r}
# TODO:
dplyr::filter(big_data, Year %in% ww_I) -> sub_sample
sub_sample<-dplyr::mutate(sub_sample,Year=as.factor(Year))
ggplot(sub_sample, mapping= aes(colour = Year,x=Age,y=mx)) +   
geom_point()+ scale_x_discrete(breaks=big_data$Age[seq(1,111,by=10)])+ scale_y_log10()+facet_wrap(~ Sex)
summary(big_data$Age)
levels(big_data$Age)
nlevels(big_data$Age)

```

Map `Sex` to some aesthetics and facet by `Year`.

```{r}
# TODO:
dplyr::filter(big_data, Year %in% ww_I) -> sub_sample
sub_sample<-dplyr::mutate(sub_sample,Year=as.factor(Year))
ggplot(sub_sample, mapping= aes(colour = Sex,x=Age,y=mx)) +   
geom_point()+ scale_x_discrete(breaks=sub_sample$Age[seq(1,111,by=10)])+ scale_y_log10()+facet_wrap(~ Year)

```

List puzzling features on those plots.
: TODO:
A lot of people die at around 20-30 (men especially) because of the war

Do you have clues about those puzzling aspects?
Died in the second world war

TODO:# Trends## Infant mortality ratePlot mortality quotients at ages $0, 1, 5$  as a function of time.
Distinguish women and men.

```{r}
dplyr::filter(big_data, Age %in% c(0,1,5)) -> infant
glimpse(big_data)
p<-ggplot(mapping= aes(colour = Sex,x=Year,y=mx)) +   
  geom_point() +
  facet_wrap(~ Age)
p%+%infant
# TODO:
```

Can you match abrupt changes and historical events?
: TODO:
The spanish flu outbreak of 1918 killed a lot of people and infants. 

## As times goes byPlot mortality quotients at ages $15, 20, 40, 60$  as a function of time.
Distinguish women and men.

Note again that you may tweak the `ggplot` object you just created and feed it with new data
using `%+%` from `ggplot`.

```{r}
ages <- c(15, 20, 40, 60)
dplyr::filter(big_data, Age %in% ages) -> infant2
p%+%infant2
# TODO:
```

TODO:
: List puzzling features
In some years the mortality quotient increases abruptly. This is due to wars, famine and disease outbreaks. 
## Mortality quotients

  - Plot mortality quotient as a function of age for years $1825, 1850, 1875, \ldots, 2000$, facet with respect to `Sex`
- Compare different scales (logarithmic or linear) on the different axes

```{r}
years <- seq(1825, 2000, by = 25)
dplyr::filter(big_data, Year %in% years) -> mq
ggplot(mq, mapping= aes(x=Age,y=mx)) +   
  geom_point(aes(colour=Year)) +scale_x_discrete(breaks=sub_sample$Age[seq(1,111,by=10)])+ scale_y_log10()+
  facet_wrap(~ Sex)

# TODO:
```

# Rearrangement

From data.frame `data_m, data_f`, extract a numerical matrix `A` where rows correspond to year (say `t`), columns to age (say `x`), and entry
for year `t` and age `x` is the mortality quotient at age `x` for year `t`. Finally, we take logarithms:  $A_{t,x} = \log m_{t,x}$.

You may use functions `pivot_wider`, `pivot_longer` from `tidyr::` package.

```{r}
# TODO
pivot<- function(data){
  data%>%
  select(Age,mx,Year) %>% tidyr::pivot_wider(names_from = Age, values_from = mx) %>%
  column_to_rownames(var = "Year") -> pivot1
  pivot1=data.matrix(pivot1)
  pivot1
  }

pivot(data_f) -> pivot1



```

> Factorizations of $A$ play a significant role in demographic modeling. More on this topic in PCA chapter.

# Life expectancy

Write a function that takes as input a vector of mortality quotients, as well as an age, and returns the residual life expectancy corresponding to the vector and the given age.

```{r}
# TODO:
pivot1["2017",]->x

rex<- function(x,y){
  sum(cumprod(1-x[(y+1):length(x)]))
}

rex(x, 0)

```


Write a function that takes as input
a _life table_ and returns a matrix $E$ where rows correspond to `Year`
and columns to `Age` with $E_{t,x}$ equal to the residual life expectancy at age $x$ for year $t$.

```{r}
# TODO:
full_rex <- function(lt)
{
  # TODO:
  retype(lt)->lt
  pivot(lt)->pivottable
  full <- pivottable
  for (Year in 1:nrow(pivottable)){
    for(Age in 1:ncol(pivottable)){
      full[Year,Age]=rex(pivottable[Year,],Age)
    }
  }
  full
  
  
}
lt <- readr::read_delim(paste(file_path,sep=""), delim = " ")

```

Plot residual life expectancy as a function of `Year` at ages $60$ and $65$.
You may proceed by first performing aggregation on `data_...` (grouping by `Year`),
and then compute residual life expectancy at different ages
by slicing the `mx` column. Recall that when mapping hazard rate to
residual life expextancy, the order of multiplications matters. How could you secure
your computation so that it still works if the rows of `data_...` are shuffled ?

Function  `tidyr::pivot_longer` may help.

```{r}
# TODO:
full_rex(lt) -> a
a=a[,c(61,66)]
a=data.frame(a)
rownames_to_column(a, var ="Year") -> a
pivot_longer(a,-Year,names_to = "Age", values_to= "rex")->a

a %>% ggplot(mapping= aes(x=Year,y=rex)) +   
  geom_point() +scale_x_discrete(breaks=a$Year[seq(0,length(a$Year),by=100)])+ scale_y_log10()+
  facet_wrap(~ Age)

```

# References

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
